---
title: "Rebilance Microbiome Report"
output:
  prettydoc::html_pretty:
    theme: cayman
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Welcome to your Rebilance report. Lets begin. 

Hi John, 

**Your gut microbiota has been found to be more unhealthy than healthy compared to our reference database. Not to worry - we're on it!**

```{r, echo = FALSE, message = FALSE}
#loading relevent libraries. 
library("mixOmics")
library("readr")
library(tidyverse)
library(pROC)
library(caret)


# We can load in our model
splsda <- readRDS("splsda_model.rds")

# And also our test set of HFM individuals. 

testing_set <- readRDS("testing_set.rds")

# Initialize a list to store the dataframes
df_list <- list()

# Split the testing_set into four dataframes
for (i in 1:nrow(testing_set)) {
  df_list[[i]] <- testing_set[i, ]
}

user <- df_list[[3]]

user_score <- user

user <- user[, -c(1:2)]

dfm <- user

new_sample_scores <- predict(splsda, newdata = dfm, type = "scores")


# Assuming df1 is a dataframe with a single row (sample)
# new_sample_scores <- predict(splsda, newdata = df1, type = "scores")  # Uncomment this line if splsda and df1 are defined

# Extract the predicted score for the first component and first class
predicted_score <- new_sample_scores$predict[, , 1]

# Calculate the percentages
healthy_percentage <- round(predicted_score[1] * 100)
unhealthy_percentage <- round(predicted_score[2] * 100)
```


```{r, results='asis'}
# Print the formatted message
cat("Our model predicts that there is a ", unhealthy_percentage, "% chance that you have an unhealthy microbiome, and only a ", healthy_percentage, "% chance that you have a healthy one.")

```

**You may be wondering, how can we help you? What we have done is run your data against our model, and have created a curated list of microbial supplements that will help correct your gut microbiome.**

Some of these microbes inlcude the following:

```{r, echo=FALSE, results='asis'}
dft <- dfm

df_long <- readRDS("joined_data")
#df_long <- df_long[, -1]


# Step 1: Find the top 30 positive contributing genera for healthy users
get_top_healthy_genera <- function(n_genera = 30){
    # Get the loadings of the genera in the first component
    loadings <- splsda$loadings$X[, "comp1"]

    # Sort the loadings in decreasing order
    sorted_loadings <- sort(loadings, decreasing = TRUE)

    # Filter the loadings to keep only the negative ones (associated with 'healthy')
    healthy_loadings <- sorted_loadings[sorted_loadings < 0]

    # Get the names of the top contributing genera for healthy individuals
    top_healthy_genera <- names(healthy_loadings)[1:n_genera]
    
    return(top_healthy_genera)
}
top_healthy_genera <- get_top_healthy_genera()

# Step 2: Filter out the df_long dataframe to include only these genera for healthy and unhealthy users
df_long_filtered <- df_long[, c("Group", top_healthy_genera)]

# Step 3: Calculate the average abundances of each of these genera in the df_long dataframe
average_abundance <- colMeans(df_long_filtered[df_long_filtered$Group == 0, -1])

# Step 4: Perform steps 2 and 3 on the dft dataframe
# The dft dataframe contains only genera columns (and no 'Group' column)
dft_filtered <- dft[, top_healthy_genera]
average_abundance_dft <- colMeans(dft_filtered)

# Step 5: If the average abundance for any of these genera is higher in the average healthy abundances than they are in the dft dataframe, print them out.
deficient_genera <- names(average_abundance)[average_abundance > average_abundance_dft]

# Calculate the difference in abundance for deficient genera
abundance_difference <- average_abundance[deficient_genera] - average_abundance_dft[deficient_genera]

# Create a character vector of the top 5 deficient genera names
top_5_deficient_names <- head(deficient_genera, 5)

# Create a sequence of numbers from 1 to 5
numbers <- seq_along(top_5_deficient_names)

# Combine the numbers and names with formatting
formatted_genera <- sprintf("%d. *%s*", numbers, top_5_deficient_names)

# Create a markdown list by concatenating the formatted items with line breaks
formatted_list <- paste0("* ", formatted_genera, collapse = "\n")

# Print the formatted list
cat(formatted_list, "\n")

```

When you recieve your probiotic kit, a more detailed list of its contents will be provided. 

You may be wondering how your microbiome compares to our model. To communicate your microbial health, we have developed a score to help you keep track of your microbiome. This score is a representation of your level of "dysbiosis" compared to our model.

```{r, echo=FALSE,message=FALSE, warning=FALSE}
# Load necessary packages
library(dplyr)
library(ggplot2)
library(tidyr)

# Define the function to get top genera
get_top_genera <- function(splsd, n_genera = 30){
    loadings <- splsda$loadings$X[, "comp1"]
    sorted_loadings <- sort(loadings, decreasing = TRUE)
    top_genera <- names(sorted_loadings)[1:n_genera]
    return(top_genera)
}

# Get the top 30 genera
top_genera <- get_top_genera(splsda)

# Filter df_long by these genera
df_long_filtered <- df_long %>% select(SampleID, Group, top_genera)

# Get loadings for these genera
loadings <- splsda$loadings$X[top_genera, "comp1"]

# Initialize a new dataframe to hold the multiplied values
df_long_filtered_multiplied <- df_long_filtered[, c("SampleID", "Group")]

# Multiply each column by its loading
for (genus in top_genera) {
  df_long_filtered_multiplied[[genus]] <- df_long_filtered[[genus]] * loadings[genus]
}

# Gather the data into a long format for plotting
df_long_filtered_multiplied <- df_long_filtered_multiplied %>% 
                    pivot_longer(cols = top_genera, 
                                 names_to = "Genus", 
                                 values_to = "Score")

# Calculate average scores for each group
average_scores <- df_long_filtered_multiplied %>%
                    group_by(Group) %>%
                    summarise(Avg_Score = mean(Score), .groups = 'drop')

# Repeat the process for the new user data
user_score_filtered <- user_score %>% select(SampleID, Group, top_genera)
user_score_filtered_multiplied <- user_score_filtered[, c("SampleID", "Group")]

for (genus in top_genera) {
  user_score_filtered_multiplied[[genus]] <- user_score_filtered[[genus]] * loadings[genus]
}

user_score_long_filtered_multiplied <- user_score_filtered_multiplied %>% 
                    pivot_longer(cols = top_genera, 
                                 names_to = "Genus", 
                                 values_to = "Score")

user_score_long_filtered_multiplied$Group <- "New User"

# Calculate average of summed scores for the new user
avg_sum_scores_user_score <- user_score_long_filtered_multiplied %>%
  group_by(SampleID, Group) %>%
  summarise(Avg_Score = mean(Score), .groups = 'drop')

# Convert Group to character for both dataframes
average_scores$Group <- ifelse(average_scores$Group == 0, "Healthy", "Unhealthy")
avg_sum_scores_user_score$Group <- "New User"

# Combine the average scores for each group with the new user's score
all_scores <- bind_rows(average_scores, avg_sum_scores_user_score)

# Convert Group to factor for clearer plot labels
all_scores$Group <- factor(all_scores$Group,
                           levels = c("Healthy", "Unhealthy", "New User"))

# Plot the distribution of scores for each group
barplot <- ggplot(all_scores, aes(x = Group, y = Avg_Score, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Average Scores by Health Category",
       x = "Category", y = "Dysbiosis Score") +
  scale_fill_manual(values = c("green", "red", "orange"))
```

```{r, results='asis'}
# Extract the health score for the new user
health_score <- sprintf("%.3f", all_scores$Avg_Score[all_scores$Group == "New User"])

# Extract the average scores for healthy and unhealthy folks
avg_score_healthy <- sprintf("%.3f", all_scores$Avg_Score[all_scores$Group == "Healthy"])
avg_score_unhealthy <- sprintf("%.3f", all_scores$Avg_Score[all_scores$Group == "Unhealthy"])

# Print the formatted message
cat("Our model gives you a dysbiosis score of ", health_score, ". Our model's average score for healthy folks is ", avg_score_healthy, ", and for unhealthy folks is ", avg_score_unhealthy, ".")

```

Here we can visualize this difference with a bar graph:

```{r}
print(barplot)
```

*If your score looks much higher than the unhealthy score - don't worry, these scores are averages, and being higher than the average unhealthy score is not nessesarily a serious problem. *

**However, it is not ideal**

To get you back on track, we have performed an analysis and have prepared a personal probiotic formula to get you back on track. 

You will receive a shipment with our probiotics in a few days. If you have any questions, please feel free to reach out. 

Best regards, 

***Rebilance team***



