---
title: "Rebilance Microbiome Report"
output:
  prettydoc::html_pretty:
    theme: cayman
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Welcome to your Rebilance report. Lets begin. 

Hi John, 

**Your gut microbiota has been found to be more unhealthy than healthy compared to our reference database. Not to worry - we're on it!**

```{r, echo = FALSE, message = FALSE, warning=FALSE}
#loading relevent libraries. 
library("mixOmics")
library("readr")
library(tidyverse)
library(pROC)
library(caret)
library(tm)
library(wordcloud)
library(RColorBrewer)

# We can load in our model
splsda <- readRDS("splsda_model.rds")

# And also our test set of HFM individuals. 

testing_set <- readRDS("testing_set.rds")

# Initialize a list to store the dataframes
df_list <- list()

# Split the testing_set into four dataframes
for (i in 1:nrow(testing_set)) {
  df_list[[i]] <- testing_set[i, ]
}

### TO CHANGE USER, REPLACE NUMBER BELOW (1,2,3)
user <- df_list[[3]]
## ^^^
user_score <- user

user <- user[, -c(1:2)]

dfm <- user

new_sample_scores <- predict(splsda, newdata = dfm, type = "scores")


# Assuming df1 is a dataframe with a single row (sample)

# Extract the predicted score for the first component and first class
predicted_score <- new_sample_scores$predict[, , 1]

# Calculate the percentages
healthy_percentage <- round(predicted_score[1] * 100)
unhealthy_percentage <- round(predicted_score[2] * 100)
```


let's begin by introducing you to your gut bacteria: here are some of the major players in your gut:

```{r, echo=FALSE, fig.width=4, fig.height=4, fig.align='center'}
# Save the column names (genera)
genera <- colnames(user)

# Transpose the data frame and convert to numeric
data <- suppressWarnings(as.numeric(as.character(unlist(user))))

# Assign the genera as the names of the vector
names(data) <- genera

# Remove missing values
data <- data[!is.na(data)]

# Sort by abundance and select the top 15 genera
data <- sort(data, decreasing = TRUE)
data <- data[1:15]

# Create the word cloud with all words displayed horizontally
wordcloud(words = names(data), freq = data, min.freq = 1,
          max.words=200, random.order=FALSE, rot.per=0, 
          colors=brewer.pal(8, "Dark2"))
```


```{r, results='asis', warning=FALSE}
# Print the formatted message
cat("Our model predicts that there is a ", unhealthy_percentage, "% chance that you have an unhealthy microbiome, and only a ", healthy_percentage, "% chance that you have a healthy one.")

```

**You may be wondering, how can we help you? What we have done is run your data against our model, and have created a curated list of microbial supplements that will help correct your gut microbiome.**

Some of these microbes inlcude the following:

```{r, echo=FALSE, results='asis', warning=FALSE}
dft <- dfm

df_long <- readRDS("joined_data")
#df_long <- df_long[, -1]

### This is the same script as before, with little modifications to print out the needed genera


# Step 1: Find the top 30 positive contributing genera for healthy users
get_top_healthy_genera <- function(n_genera = 30){
    # Get the loadings of the genera in the first component
    loadings <- splsda$loadings$X[, "comp1"]

    # Get the absolute values of the loadings
    abs_loadings <- abs(loadings)

    # Sort the absolute loadings in decreasing order
    sorted_abs_loadings <- sort(abs_loadings, decreasing = TRUE)

    # Filter the loadings to keep only the original negative ones (associated with 'healthy')
    healthy_loadings <- sorted_abs_loadings[loadings < 0]

# Get the names of the top contributing genera for healthy individuals
top_healthy_genera <- names(healthy_loadings)[1:n_genera]
    
    return(top_healthy_genera)
}
top_healthy_genera <- get_top_healthy_genera()

# Step 2: Filter out the df_long dataframe to include only these genera for healthy and unhealthy users
df_long_filtered <- df_long[, c("Group", top_healthy_genera)]

# Step 3: Calculate the average abundances of each of these genera in the df_long dataframe
average_abundance <- colMeans(df_long_filtered[df_long_filtered$Group == 0, -1])

# Step 4: Perform steps 2 and 3 on the dft dataframe
# The dft dataframe contains only genera columns (and no 'Group' column)
dft_filtered <- dft[, top_healthy_genera]
average_abundance_dft <- colMeans(dft_filtered)

# Step 5: If the average abundance for any of these genera is higher in the average healthy abundances than they are in the dft dataframe, print them out.
deficient_genera <- names(average_abundance)[average_abundance > average_abundance_dft]

# Calculate the difference in abundance for deficient genera
abundance_difference <- average_abundance[deficient_genera] - average_abundance_dft[deficient_genera]

# Create a character vector of the top 5 deficient genera names
top_5_deficient_names <- head(deficient_genera, 5)

# Create a sequence of numbers from 1 to 5
numbers <- seq_along(top_5_deficient_names)

# Combine the numbers and names with formatting
formatted_genera <- sprintf("%d. *%s*", numbers, top_5_deficient_names)

# Create a markdown list by concatenating the formatted items with line breaks
formatted_list <- paste0("* ", formatted_genera, collapse = "\n")

# Print the formatted list
cat(formatted_list, "\n")

```

When you receive your probiotic kit, a more detailed list of its contents will be provided. 

You may be wondering how your microbiome compares to our model. To communicate your microbial health, we have developed a score to help you keep track of your microbiome. This score is a representation of your level of "dysbiosis" compared to our model.

```{r, warning=FALSE, results='asis'}
# Load necessary packages
library(dplyr)
library(ggplot2)
library(tidyr)

## The following script is used to calculate and visually communicate a "dysbiosis" score to the user. We take the average abundances of each genera for healthy and unhealthy users and multiply by the associated loading value as gained from the model. We do the same for the new user, and then sum both data sets (first as average, new user is just one so its not averaged) We then transform the data to only have positive values - this is done only to make it easier to visually communicate to the user the difference in dysbiosis scores. We then plot and print out an automated message. 

# Get all genera
all_genera <- colnames(df_long)[-(1:2)]  # assuming the first two columns are SampleID and Group

# Filter df_long by these genera
df_long_filtered <- df_long %>% select(SampleID, Group, all_genera)

# Get loadings for these genera
loadings <- splsda$loadings$X[all_genera, "comp1"]

# Initialize a new dataframe to hold the multiplied values
df_long_filtered_multiplied <- df_long_filtered[, c("SampleID", "Group")]

# Multiply each column by its loading
for (genus in all_genera) {
  df_long_filtered_multiplied[[genus]] <- df_long_filtered[[genus]] * loadings[genus]
}

# Gather the data into a long format for plotting
df_long_filtered_multiplied <- df_long_filtered_multiplied %>% 
                    pivot_longer(cols = all_genera, 
                                 names_to = "Genus", 
                                 values_to = "Score")

# Calculate average scores for each group
average_scores <- df_long_filtered_multiplied %>%
                    group_by(Group) %>%
                    summarise(Avg_Score = mean(Score), .groups = 'drop')

# Repeat the process for the new user data
user_score_filtered <- user_score %>% select(SampleID, Group, all_genera)
user_score_filtered_multiplied <- user_score_filtered[, c("SampleID", "Group")]

for (genus in all_genera) {
  user_score_filtered_multiplied[[genus]] <- user_score_filtered[[genus]] * loadings[genus]
}

user_score_long_filtered_multiplied <- user_score_filtered_multiplied %>% 
                    pivot_longer(cols = all_genera, 
                                 names_to = "Genus", 
                                 values_to = "Score")

user_score_long_filtered_multiplied$Group <- "New User"

# Calculate average of summed scores for the new user
avg_sum_scores_user_score <- user_score_long_filtered_multiplied %>%
  group_by(SampleID, Group) %>%
  summarise(Avg_Score = mean(Score), .groups = 'drop')

# Convert Group to character for both dataframes
average_scores$Group <- ifelse(average_scores$Group == 0, "Healthy", "Unhealthy")
avg_sum_scores_user_score$Group <- "New User"

# Combine the average scores for each group with the new user's score
all_scores <- bind_rows(average_scores, avg_sum_scores_user_score)

# Calculate the minimum score across all groups
min_score <- min(all_scores$Avg_Score)

# Calculate the shift needed to make the minimum score zero (or a small positive number)
shift <- abs(min_score) + 0.01  # add a small constant to ensure all scores are positive

# Shift all the scores
all_scores$Avg_Score <- all_scores$Avg_Score + shift

# Convert Group to factor for clearer plot labels
all_scores$Group <- factor(all_scores$Group,
                           levels = c("Healthy", "Unhealthy", "New User"))

# Plot the distribution of scores for each group
barplot <- ggplot(all_scores, aes(x = Group, y = Avg_Score, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Average Scores by Health Category",
       x = "Category", y = "Dysbiosis Score") +
  scale_fill_manual(values = c("Healthy" = "green", "Unhealthy" = "red", "New User" = "orange"))

# Extract the health score for the new user
health_score <- sprintf("%.3f", all_scores$Avg_Score[all_scores$Group == "New User"])

# Extract the average scores for healthy and unhealthy folks
avg_score_healthy <- sprintf("%.3f", all_scores$Avg_Score[all_scores$Group == "Healthy"])
avg_score_unhealthy <- sprintf("%.3f", all_scores$Avg_Score[all_scores$Group == "Unhealthy"])

# Print the formatted message
cat("Our model gives you a dysbiosis score of ", health_score, ". Our model's average score for healthy folks is ", avg_score_healthy, ", and for unhealthy folks is ", avg_score_unhealthy, ".")


```

Here we can visualize this difference with a bar graph:

```{r}
print(barplot)
```

*If your score looks much higher than the unhealthy score - don't worry, these scores are averages, and being higher than the average unhealthy score is not nessesarily a serious problem. *

**However, it is not ideal**

To get you back on track, we have performed an analysis and have prepared a personal probiotic formula to get you back on track. 

You will receive a shipment with our probiotics in a few days. If you have any questions, please feel free to reach out. 

Best regards, 

***Rebilance team***



